---
title: "Matching"
author: "Linh Tran"
date: "5/18/2021"
output: html_document
---

```{r setup, include=FALSE}
library(MatchIt)
library(optmatch)
```

# `MatchIt`

`MatchIt` provides a simple and straightforward interface to various methods of matching for covariate balance in observational studies. Several matching methods are available: nearest neighbor matching, optimal pair matching, optimal full matching, genetic matching, exact matching, subclassification,.... Here is an example of the use of `MatchIt` to perform full matching and assess balance

```{r}
data("lalonde")
head(lalonde)

# full matching on the PS
m.out <- matchit(treat ~., data = lalonde, method = "full")

m.out
summary(m.out) #check covariate balance for the original and matched samples
```

* We have:
   + balance for the original sample
   + balance in the matched sample
   + percent reduction in imbalance 
   + sample size before and after matching
Smaller values for the balance statistics indicate better balance.\

We can plot the standardized mean differences in a Love plot for a visual display of balance across sample:

```{r}
# plot balance
plot(summary(m.out))
```

